---
title: "Elliptic moments for circular data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elliptic moments for circular data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ellstat)
```

## The idea

Trigonometric moments summarise a circular distribution by averaging
`cos(p*theta)` and `sin(p*theta)`. They characterise the law and they are the
natural target of the method of moments, but they are a single fixed system:
the basis cannot be adapted to the problem.

The Jacobi elliptic functions supply the missing degree of freedom. For a
modulus `k` in `[0, 1)` they deform the sine and cosine, reduce to them exactly
at `k = 0`, and keep the two properties that make the trigonometric basis
usable: periodicity, and the Pythagorean identity. Running them on a clock
rescaled by the quarter period gives a basis on the circle for every modulus.

```{r}
theta <- seq(0, 2 * pi, length.out = 7)
# at modulus zero the basis IS the trigonometric one
max(abs(ell_basis(theta, 1L, k = 0)$C[1, ] - cos(theta)))
```

## Exact finite-sample theory

Because `cn^2 + sn^2 = 1`, each basis pair is a unit vector, so the sample
elliptic moment is a mean of unit vectors. Two consequences hold for *every*
sample size, not asymptotically: the sample moment is unbiased, and its total
variance is exactly `(1 - |E_p|^2)/n`.

```{r}
theta <- rellvm(200, mu = 1, kappa = 3, k = 0.9, seed = 1)
mv <- ell_moment_var(theta, P = 3L, k = 0.9)
mv
# the realised and exact columns agree identically: this is an algebraic
# identity, not a numerical approximation
max(abs(mv[, "realised"] - mv[, "exact"]))
```

The joint covariance of several moments is exact in the same sense, where the
corresponding result for L-moments is a limit.

```{r}
J <- ell_joint_sample(theta, P = 2L, k = 0.9)
round(J$Sigma, 4)
```

## Exact inference

The circular uniform law is fully specified, so it can be simulated directly and
the rank p-value is exact for every sample size and every number of replicates.

```{r}
unif <- rellvm(60, mu = 0, kappa = 0, k = 0.9, seed = 3)
ell_exact_uniform(unif, k = 0.9, B = 199L, seed = 1L)
```

Testing several moments jointly costs power against a departure concentrated in
the first, and gains it against an antipodally symmetric departure, to which a
first-moment test is permanently blind.

```{r}
ell_joint_uniform(unif, P = 2L, k = 0.9, B = 199L, seed = 1L)
```

Confidence intervals come from inverting the test. The null here is *composite*,
the statistic depending on the mean direction as well as the concentration, so
the p-value is maximised over a set of mean directions; the interval is then
conservative rather than exact, and valid at any sample size.

```{r}
th <- rellvm(80, mu = 1, kappa = 3, k = 0.9, seed = 11)
ci <- ell_exact_ci(th, k = 0.9, kgrid = seq(0.5, 8, length.out = 20L),
                   B = 99L, level = 0.05, nmu = 6L)
c(lower = ci$lower, upper = ci$upper)
```

## Which modulus?

The efficiency of a moment estimator relative to maximum likelihood is exactly
the squared correlation between the statistic and the score, so the best modulus
maximises that correlation and can be computed without simulating anything.

```{r}
# the von Mises law is fully efficient at modulus zero and nowhere else
round(ell_catalogue("vonmises", p1 = 2, kgrid = c(0, 0.9, 0.99))$efficiency, 4)
```

The gains are largest for heavy tails. For a wrapped Cauchy law the
trigonometric moment retains about a fifth of the available information and the
elliptic moment about four fifths.

```{r}
wc <- ell_catalogue("wrappedcauchy", p1 = 0.9, kgrid = c(0, 0.9, 0.999))
round(wc$efficiency, 3)
round(c(optimal_modulus = wc$kopt, efficiency = wc$eff_kopt), 4)
```

The criterion appears to need the answer before it can be used, since the score
is unknown. It does not, because the surface is flat near its maximum: being in
the right region is nearly as good as being at the optimum.

## Where it does not help

For the von Mises, wrapped normal and cardioid families the optimal modulus is
zero and the deformation gains nothing. That is reported rather than concealed,
and is what makes the rest of the catalogue trustworthy.

```{r}
round(ell_catalogue("wrappednormal", p1 = 0.6, kgrid = c(0, 0.9))$efficiency, 3)
```
